Blum axioms and nondeterministic computation of functions

### 2016-04-17 paper.tex
paper.tex: Rewrite the explanation about Hopcroft-Ullman's definition.

This formalizes the notion of complexity class.
There are some important theorems concerning complexity classes
under the Blum axioms;
we mention only the Union Theorem \cite[p.~234]{Kozen2006}.
It states that,
if $\{f_i\}$ is any recursive list of increasing functions
such that $f_i(n) \leq f_{i+1}(n)$ for all $x$
(that is, each $f_i$ is greater than the previous),
then there is some function $g$ such that
\begin{equation*}
    \mathcal C_\Phi(g) = \bigcup_{i \in \mathbb N} \mathcal C_\Phi(f_i).
\end{equation*}
That is, the class $\mathcal C_\Phi(g)$ contains \emph{exactly}
all functions present in the classes $\mathcal C_\Phi(f_i)$.
Choosing $\Phi$ as the time complexity and $f_i(n) = n^i$
(the polynomial functions),
the union in the right is exactly the class $\P$,
the problems solvable in polynomial time.
By the Union Theorem, $\P$ is $\mathcal C_\Phi(g)$
for some recursive function $g$,
so even though $\P$ does not have an easily specifiable bounding function
such function exists nevertheless.

@@

In their 1979 book,
Hopcroft and Ullman note the need
for defining how a nondeterministic Turing machine can compute a function;
they mention one possible approach:
\begin{quotation}
    We may establish that nondeterministic time and space
    satisfy the axioms if we make an intelligent definition of what it means
    for an NTM \emph{[nondeterministic Turing machine]} to compute a function.
    For example, we might say that $\phi_i(n) = j$
    if and only if there is some sequence of choices by $M_i$ with input $0^n$
    that halts with $0^j$ on the tape,
    and no sequence of choices that leads to halting with some $0^k$, $k \neq j$,
    on the tape. \cite[p.~313]{HopcroftUllman1979}
\end{quotation}
Hopcroft and Ullman make this definition in the context of integer functions;
in our notation, this means that $\phi_w(x) = y$
if and only if there is some sequence of choices of the machine represented by $w$
when processing the input $x$ that halts with $y$ in the tape,
and no other sequence of choices that leads to halting does so with some $z$,
$z \neq y$, on the tape.

@@

\begin{proof}
    Notice that the domain in which $f$ is defined
    is the complement under $\Sigma^*$ of the halting problem.
    Therefore,
    we can pair a machine that computes $f$
    with a machine that recognizes (but not decides) the halting problem
    to actually decide the halting problem,
    as follows:

    Let $M$ be the machine that computes $f$,
    and $U$ be the universal Turing machine
    (that, on input $(w, x)$,
    either returns $\pi_w(x)$ if it exists
    or loops forever if it doesn't).
    The machine $M'$ that will solve the halting problem
    will simulate $M$ and $U$ simultaneously,
    one move for each machine.
    Since the domains of $M$ and $U$ are complementary,
    for any input $(w, x)$, exactly one of them will halt.
    If is $U$, then we accept because $w$ halts on $x$;
    otherwise, if it is $M$, we reject because $w$ never halts on $x$.
    This solves the halting problem --- a contradiction.
    Hence, no Turing machine can compute $f$.

    But $f$ can be computed under Hopcroft and Ullman's definition as follows:
    on input $(w, x)$,
    create two branches of computation.
    On the first, write $1$ on the tape and halt.
    On the second, simulate the universal Turing machine $U$ in the input,
    and if $U$ halts, write $0$ on the tape and also halt.

@@

    Thus, this machine can, under Hopcroft and Ullman's definition,
    compute the noncomputable function $f$.
\end{proof}

Therefore,
using Hopcroft and Ullman's definition,
we could enumerate functions that are not computable;
so, the image of the corresponding ``Gödel numbering''
would include not only computable functions.

>>>>

This formalizes the notion of complexity class.
There are some important theorems concerning complexity classes
under the Blum axioms;
we mention only the Union Theorem \cite[p.~234]{Kozen2006}.
It states that,
if $\{f_i\}$ is any recursive list of increasing functions
such that $f_i(n) \leq f_{i+1}(n)$ for all $x$
(that is, each $f_i$ is greater than the previous),
then there is some function $g$ such that
\begin{equation*}
    \mathcal C_\Phi(g) = \bigcup_{i \in \mathbb N} \mathcal C_\Phi(f_i).
\end{equation*}
That is, the class $\mathcal C_\Phi(g)$ contains \emph{exactly}
all functions present in the classes $\mathcal C_\Phi(f_i)$.
Choosing $\Phi$ as the time complexity and $f_i(n) = n^i$
(the polynomial functions),
the union in the right is exactly the class $\P$,
the problems solvable in polynomial time.
By the Union Theorem, $\P$ is $\mathcal C_\Phi(g)$
for some recursive function $g$,
so, even though $\P$ does not have an easily specifiable bounding function,
such function exists nevertheless.

@@

\begin{definition}[Hopcroft and Ullman's definition\footnotemark]
    \footnotetext{
        Hopcroft and Ullman's original definition \cite[p.~313]{HopcroftUllman1979}
        was defined in the context of computation of integer functions.
        We are rephrasing here in terms of strings,
        but keeping the relation they imposed on the execution branches.
    }
    If $w$ is an encoding for the Turing machine $M$,
    we say that $\phi_w(x) = y$ if and only if,
    when processing $x$,
    there is some branch of $M$ that halts with $y$ in the tape,
    and there is no branch that halts with some $z \neq x$ in the tape.
\end{definition}

@@

\begin{proof}
    Note that the domain in which $f$ is defined is exactly
    the complement (under $\Sigma^*$) of the halting problem.
    If $f$ were computable,
    there would be a deterministic machine that could compute it;
    we could, then, modify the machine to accept all inputs in the domain of $f$
    (all inputs that have some output through $f$),
    thus recognizing (albeit not deciding) the complement of the halting problem.
    But, as the halting problem itself is computable but not decidable,
    its complement cannot also be --- otherwise both would be decidable,
    a contradiction.

    But $f$ can be computed under Hopcroft and Ullman's definition as follows:
    on input $(w, x)$,
    create two branches of computation.
    On the first, write $1$ on the tape and halt.
    On the second, simulate the universal Turing machine $U$ in the input,
    and if $U$ halts, write $0$ on the tape and halt too.

@@

    Thus, under Hopcroft and Ullman's definition,
    this machine computes the function $f$,
    which we already shown to be noncomputable.
\end{proof}

Therefore,
using Hopcroft and Ullman's definition,
we could enumerate functions that are not computable;
so, the image of the corresponding ``Gödel numbering''
would include noncomputable functions.
